# Protocol

## Protocols for selection
1. Constructing genetic relationship matrix $\mathbf{G}$. The difference between a random mating population and inbred population is the way $\mathbf{G}$ is constructed. So for a pair of individuals $i$ and $j$, 
$$G_{ij}=\frac{1}{\tilde{m}}\sum_l^{\tilde{m}}\frac{(x_{il}-2p_l)(x_{jl}-2p_l)}{2(1+F)p_iq_i}$$
and $F=1$ for inbred lines. Please refer to Chapter \@basis.

2. Conducting eigenanalysis for $\mathbf{G}$.

3. Linear regression analysis for each SNP.

### Rscript pipeline for random mating population
```{r, euro-demo, collapse=TRUE}
plink2='/Users/gc5k/bin/plink_mac/plink'
dat="./data/euro/euro"

#make-grm
grmCmd=paste(plink2, "--bfile ", dat, "--make-grm-gz --out ", dat)
system(grmCmd)

gz=gzfile(paste0(dat, ".grm.gz"))
grm=read.table(gz, as.is = T)
Ne=-1/mean(grm[grm[,1]!=grm[,2], 4])
Me=1/var(grm[grm[,1]!=grm[,2], 4])
print(paste("Ne=", format(Ne, digits = 2), "Me=", format(Me, digits = 2)))

#pca
pcaCmd=paste(plink2, "--bfile ", dat, "--pca 5 --out ", dat)
system(pcaCmd)
barplot(read.table(paste0(dat, ".eigenval"), as.is = T)[,1]/2, border = F)

pc=read.table(paste0(dat, ".eigenvec"), as.is = T)
plot(pc[,3], pc[,4], xlab="Eigenvector 1", ylab="Eigenvector 2", bty="n", main="Eigenspace", bty="n", col=ifelse(pc[,3]>0, "red", "blue"), pch=16, cex=0.5)
#make-grm
source("~/R/MyLib/manhattan.R")
liCmd=paste0(plink2, " --linear --bfile ", dat, " --pheno ", dat, ".eigenvec --out ", dat)
system(liCmd)

#plot
EigenRes=read.table(paste0(dat, ".assoc.linear"), as.is = T, header = T)
EigenRes$Praw=EigenRes$P
gc=qchisq(median(EigenRes$P), 1, lower.tail = F)/qchisq(0.5, 1, lower.tail = F)
print(paste("GC = ", format(gc, digits = 4)))
EigenRes$P=pchisq(qchisq(EigenRes$Praw, 1, lower.tail = F)/gc, 1, lower.tail = F)
manhattan(EigenRes, title="EigenGWAS 1", pch=16, cex=0.3, bty='n')

#QQplot
chiseq=rchisq(nrow(EigenRes), 1)
qqplot(chiseq, qchisq(EigenRes$Praw, 1, lower.tail = F), xlab=expression(paste("Theoretical ", chi[1]^2)), ylab=expression(paste("Observed ", chi[1]^2)), bty="n", col="grey", pch=16, cex=0.5)
points(sort(chiseq), sort(qchisq(EigenRes$P, 1, lower.tail = F)), col="black", pch=16, cex=0.5)
legend("topleft", legend = c("Raw", "GC correction"), pch=16, cex=0.5, col=c("grey", "black"), bty='n')
abline(a=0, b=1, col="red", lty=2)

```


### Rscript pipeline for inbred population

```{r, arab-demo, collapse=TRUE}
plink2='/Users/gc5k/bin/plink_mac/plink'
dat="./data/arab/arab"

#make-grm
grmCmd=paste(plink2, "--bfile ", dat, "--make-grm-gz --out ", dat)
system(grmCmd)

gz=gzfile(paste0(dat, ".grm.gz"))
grm=read.table(gz, as.is = T)
Ne=-1/mean(grm[grm[,1]!=grm[,2], 4]/2)
Me=1/var(grm[grm[,1]!=grm[,2], 4]/2)
print(paste("Ne=", format(Ne, digits = 2), "Me=", format(Me, digits = 2)))

#pca
pcaCmd=paste(plink2, "--bfile ", dat, "--pca 5 --out ", dat)
system(pcaCmd)
barplot(read.table(paste0(dat, ".eigenval"), as.is = T)[,1]/2, border = F)

pc=read.table(paste0(dat, ".eigenvec"), as.is = T)
plot(pc[,3], pc[,4], xlab="Eigenvector 1", ylab="Eigenvector 2", bty="n", main="Eigenspace", bty="n", col=ifelse(pc[,3]>0, "red", "blue"), pch=16, cex=0.5)
#make-grm
source("~/R/MyLib/manhattan.R")
liCmd=paste0(plink2, " --linear --bfile ", dat, " --pheno ", dat, ".eigenvec --out ", dat)
system(liCmd)

#plot
EigenRes=read.table(paste0(dat, ".assoc.linear"), as.is = T, header = T)
EigenRes$Praw=EigenRes$P
gc=qchisq(median(EigenRes$P), 1, lower.tail = F)/qchisq(0.5, 1, lower.tail = F)
print(paste("GC = ", format(gc, digits = 4)))
EigenRes$P=pchisq(qchisq(EigenRes$Praw, 1, lower.tail = F)/gc, 1, lower.tail = F)
manhattan(EigenRes, title="EigenGWAS 1", pch=16, cex=0.3, bty='n')

#QQplot
chiseq=rchisq(nrow(EigenRes), 1)
qqplot(chiseq, qchisq(EigenRes$Praw, 1, lower.tail = F), xlab=expression(paste("Theoretical ", chi[1]^2)), ylab=expression(paste("Observed ", chi[1]^2)), bty="n", col="grey", pch=16, cex=0.5)
points(sort(chiseq), sort(qchisq(EigenRes$P, 1, lower.tail = F)), col="black", pch=16, cex=0.5)
legend("topleft", legend = c("Raw", "GC correction"), pch=16, cex=0.5, col=c("grey", "black"), bty='n')
abline(a=0, b=1, col="red", lty=2)

```

## Protocol for predicted eigenvectors
The prediction accuracy can be written as
$$R^2 \approx \frac{1}{1+\frac{n_e}{m}}$$
