# Notes {#notes-1}

## APY

APY is application of [blockwise matrix inversion](https://www.schmidheiny.name/teaching/matrixalgebra2up.pdf), or [wiki link](https://en.wikipedia.org/wiki/Block_matrix)

$$
\begin{bmatrix}
P_{11} & P_{12}\\
P_{21} & P_{22}
\end{bmatrix}
^{-1} 
=
\begin{bmatrix}
P_{11}^-1+P_{11}^{-1}P_{12}F^{-1}P_{21}P_{11}^{-1}&P_{11}^{-1}P_{12}F^{-1}\\
F^{-1}P_{21}P_{11}^{-1}&F^{-1}
\end{bmatrix}$$

in which $F=P_{22}-P_{21}P_{11}^{-1}P_{12}$ is assumed nonsingular.



A much original idea of APY resembles Quaas and Pollak (J Anim Sci, 1980, 51:1277-87), an example of which can be found in Lynch&Walsh page 759-61. 

$G^{-1}=
\left[
\begin{matrix} I& -P_{cn}\\
0&I
\end{matrix}
\right]
\left[
\begin{matrix} G_{cc}^{-1}& 0\\
0&M_{nn}^{-1}
\end{matrix}
\right]
\left[
\begin{matrix} I& 0\\
-P_{nc}&I
\end{matrix}
\right]
=
\left[
\begin{matrix}G_{cc}^{-1}&-P_{cn}M_{nn}^{-1}\\
0&M^{-1}_{nn} \end{matrix}
\right]
\left[
\begin{matrix}I&0\\
-P_{nc}&I \end{matrix}
\right]$

Citation: Misztal, I, 2016, Inexpensive computation of the inverse of the genomic relationship matrix in populations with small effective population size, $Genetics$, 202:401-409.


in which $G$ is the numerical relationship matrix, and  $P_{cn}=G_{cc}^{-1}G_{cn}$ and $P_{nc}=G_{nc}G_{cc}^{-1}$. $G_{cc}$, $G_{cn}$, and $G_{nc}$ ($G_{cn}=G_{nc}$) are, "core" to "core", "core" to "non-core", and "non-core" to "core" relationship matrix.

$M_{nn}=diag(G_{nn})-diag(P_{cn}^TG_{cn})$

### A numerical example

This example is taken from appendix in  Misztal's paper.

$G=\left[ \begin{matrix}
0.81 & 0 & 0 & 0.80 & -0.80\\
 &0.81 & 0 & 0.80 &-0.80\\
 & & 0.01& 0 & 0\\
 & & & 1.61 & -1.60\\
 symm. & & & & 1.61
\end{matrix} \right]$

and $G_{cc}^{-1}=\left[ \begin{matrix} 1.235 & 0\\
0 & 1.235
\end{matrix} \right]$

$G_{cn}=G_{nc}^T=\left [
\begin{matrix}
0 & 0.80 & -0.80\\
0 & 0.80 & -0.80
\end{matrix} \right]$

and

$P_{cn}=G_{cc}^{-1}G_{cn}=\left [ \begin{matrix}
0.00 & 0.988 & -0.988\\
0.00 & 0.988 & -0.988
\end{matrix} \right ]$

$M_{nn}=diag(G_{nn})-diag(P_{cn}^TG_{cn}) = \left [ \begin{matrix}0.01 & &\\
& 1.61 & \\
& & 1.61
\end{matrix} \right]
-
\left [ \begin{matrix}0.00 & &\\
& 1.58 & \\
& & 1.58
\end{matrix} \right]=\left [ \begin{matrix}0.01 & &\\
& 0.03 & \\
& & 0.03
\end{matrix} \right]$

and 

$G^{-1}=\left[ \begin{matrix}
66.8 & 65.5 & 0 & -33.1 & 33.1\\
 &66.804 & 0 & -33.1 & 33.1\\
 & & 100& 0 & 0\\
 & & & 33.6 & 0\\
 symm. & & & & 33.6
\end{matrix} \right]$

This is APY inversed $G$, and for a comparison, a regular inverse of $G$ is quit different
$G^{-1}_{reg}=\left[ \begin{matrix}
40.6 & 39.4 & 0 & -19.9 & 19.9\\
 &40.6 & 0 & -19.9 & 19.9\\
 & & 100& 0 & 0\\
 & & & 60.0 & 39.9\\
 symm. & & & & 60.0
\end{matrix} \right]$

```{r apy}
GRM_C=matrix(c(0.81, 0, 0, 0.81), 2, 2, byrow = T)
GRM_N=matrix(0, 3, 3)
diag(GRM_N)=c(.01, 1.61, 1.61)
GRM_N[2,3]=GRM_N[3,2]=-1.6
GRM_CN=matrix(0, 2, 3)
GRM_CN[,2]=0.8
GRM_CN[,3]=-0.8

GRM=rbind(cbind(GRM_C, GRM_CN), cbind(t(GRM_CN), GRM_N))
print(GRM)
print(solve(GRM)) #original

IC=solve(GRM_C)
Pcn=IC %*% GRM_CN
Pnc=t(GRM_CN) %*% IC
Pcn_Gcn=t(Pcn) %*% GRM_CN
Mnn=diag(diag(GRM_N)-diag(Pcn_Gcn), nrow=nrow(GRM_N), ncol=nrow(GRM_N))
IMnn=solve(Mnn)

v1_1=rbind(IC, matrix(0, nrow=nrow(GRM_N), ncol=ncol(GRM_C)))
v1_2=rbind(-1*Pcn%*%IMnn, IMnn)
V1=cbind(v1_1, v1_2)

v2_1=rbind(diag(1, nrow=nrow(GRM_C), ncol=nrow(GRM_C)), -1*Pnc)
v2_2=rbind(matrix(0, nrow=nrow(GRM_C), ncol=nrow(GRM_N)), diag(1, nrow(GRM_N), ncol(GRM_N)))
V2=cbind(v2_1, v2_2)

IV=V1 %*% V2
print(IV) #apy
IVI=solve(IV)
print(IVI)
```


## LD score regression
```{r, eval=FALSE, tidy=TRUE}
source("~/R/MyLib/shotgun.R")
REP=50
N = 500
M = 1000
h2 = 0.5

h2E=array(0, dim=c(2,REP))

for(rep in 1:REP) {
  fq = runif(M, 0.05, 0.95)
  Dl = array(0, dim=M)
  Dl = runif(M, 0.8, 0.9)
  Dl[seq(10, M, 10)] = 0
  
  G = GenerateGenoDprime(fq, Dl[1:(length(Dl)-1)], N)
  b = rnorm(M, 0, sqrt(h2/M))
  bv= G%*%b
  Y = bv + rnorm(N, 0, 1)
  chi = array(0, dim=M)
  for(i in 1:M) {
    sm = summary(lm(Y~G[,i]))
    chi[i] = sm$coefficients[2,3]^2
  }
  plot(chi, b^2*fq*(1-fq), pch=16)
  lds = array(0, dim=M)
  for(i in 1:(M/10)) {
    cg = cor(G[,((i-1)*10+1):(i*10)])
    for(j in 1:10) {
      lds[(i-1)*10+j] = sum(cg[,j]^2)
    }
  }
  ldSc=lm(chi~lds)
  h2E[1,rep]=ldSc$coefficients[2]*M/N
  h2E[2,rep]=var(bv)/var(Y)
}
plot(h2E[1,], h2E[2,])

```


## Matrix inversion [wiki](https://en.wikipedia.org/wiki/Invertible_matrix#Inversion_of_2_%C3%97_2_matrices)
### 2 X 2
The cofactor equation listed above yields the following result for 2 ?? 2 matrices. Inversion of these matrices can be done as follows

$$\text{A}^{-1}=\left[ \begin{matrix} 
a & b\\
c & d
\end{matrix} \right] ^{-1}
=\frac{1}{det (\text{A})}\left[ \begin{matrix} 
A & B\\
C & D
\end{matrix} \right]
=\frac{1}{ad-bc}\left[ \begin{matrix} 
d & -b\\
-c & a
\end{matrix} \right]
$$
Where the scalar $A$ is not to be confused with the matrix A.

### 3 X 3
$$\text{A}^{-1}=\left[ \begin{matrix} 
a & b & c\\
d & e & f\\
g & h & i
\end{matrix} \right] ^{-1}
=\frac{1}{det (\text{A})}\left[ \begin{matrix} 
A & B & C\\
D & E & F\\
G & H & I
\end{matrix} \right] ^{T}
=\frac{1}{det (\text{A})}\left[ \begin{matrix} 
A & D & G\\
B & E & H\\
C & F & I
\end{matrix} \right]
$$

in which 
$$\begin{matrix}
A=(ei-fh), & D=-(bi-ch), & G = -(bf-ce), \\
B=-(di-fg), & E=(ai-cg), & H = -(af-cd), \\
C=(dh-eg), & F=-(ah-bg), & I = (ae-bd).
\end{matrix}
$$